---
title: "Ordinal Logistic Regression in R"
author: "Clay Ford, UVA Library"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


This is an R Markdown Notebook. When you execute code within the notebook, the results appear beneath the code. 

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter* (Win/Linux) or *Cmd+Shift+Return* (Mac). 

```{r}
# trees is a data set that comes with R
pairs(trees)
```

To hide the output, click the Expand/Collapse output button. To clear results (or an error), click the "x". 

You can also press *Ctrl+Enter* (Win/Linux) or *Cmd+Return* (Mac) to run one line of code at a time (instead of the entire chunk).

Add a new R code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac).  

## CODE ALONG 0

Try the `cumsum()` function with vector `1:5`

```{r}
cumsum(1:5)
```

## Load Packages

We'll use the following packages in today's workshop.

```{r}
library(vcdExtra)
library(MASS)
library(car)
library(effects)
```

