---
title: "polr simulation"
author: "Clay Ford"
date: "2022-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(MASS)
data("Arthritis", package = "vcd")
m1 <- polr(Improved ~ Treatment, data = Arthritis)
```


## Model assessment via simulation

A good model should generate data that looks similar to our observed data. We can use the `simulate()` function to rapidly simulate many new sets of our response variable. Below we use our model to generate 200 new sets of the Improved variable. The result is a data frame with 200 columns, where each column is a simulated outcome from our model. The number of rows equals the sample size of our original data frame: 84. We're taking our 84 observed predictors and using our model to simulate the outcome, Improved.

```{r}
sims <- simulate(m1, nsim = 200)
dim(sims)
```

Here's a glimpse of the first 5 simulations:

```{r}
sims[1:5,1:5]
```

Let's calculate the proportion of each response within each simulation using `xtabs()` and `proportions()`. The last function, `as.data.frame()` converts the table to a data frame and names the column containing the proportions "P" (the default is "Freq").

```{r}
xtabs(~ sims$sim_1) |>
  proportions() |>
  as.data.frame(responseName = "P")
```

To calculate proportions for _all columns_, we can convert the last chunk of code into a function and apply it to each column. Let's name it `p_df`. Notice it's the same code above with `x` in place of `sims$sim_1`.

```{r}
p_df <- function(x){
  xtabs(~ x) |> 
  proportions() |>
  as.data.frame(responseName = "P")
  }
```

Let's test it.

```{r}
p_df(sims$sim_1)
```

Now let's apply the function to every column of the sims data frame using `lapply` and save as `sims_p`. . The "L" in `lapply` means the result will be a list. 

```{r}
sims_p <- lapply(sims, p_df)
```

The result is a list with 200 data frames. Let's look at the first two using the `$` operator to extract the data frame

```{r}
sims_p$sim_1
```

```{r}
sims_p$sim_2
```

We can _combine all the data frames_ into one data frame using `rbind()`. The arguments to `rbind` are the data frames we want to combine. Instead of typing out all 200 data frames, we can use the `do.call()` function to "feed" all the data frames to the `rbind` function as follows. Notice I assign to an object called `sims_pdf`. 

```{r}
sims_pdf <- do.call(rbind, sims_p)
```

Now we have a data frame that we can use to create a plot using ggplot2.

```{r}
ggplot() +
  geom_jitter(mapping = aes(x = x, y = P), 
             data = sims_pdf, alpha = 0.25,
             width = 0.1, height = 0)
```

_How does this compare to the original data_? Let's add that to the plot. We need to create a data frame of the observed proportions. We can do that using the same code from above and name the result `obs_pdf`:


```{r}
obs_pdf <- xtabs(~ Improved, data = Arthritis) |> 
  proportions() |>
  as.data.frame(responseName = "P")
obs_pdf
```

Now update the plot to include the original data as big blue dots.

```{r}
ggplot() +
  # simulated data - gray dots
  geom_jitter(mapping = aes(x = x, y = P), 
             data = sims_pdf, alpha = 0.25,
             width = 0.1, height = 0) +
  # observed data - blue dots
  geom_point(mapping = aes(x = Improved, y = P), 
             data = obs_pdf, size = 4, color = "blue") 

```

The model doesn't seem biased in any way. It doesn't consistently over or under predict. We want the model-simulated data to hover around the observed data. However there is a good deal of variability. For example the model simulates "None" from 0.4 to 0.6. Then again this model is only using one binary predictor: Treatment. 
